SHARP II: Dark Matter Substructure, Gravitational Lens flux-ratio Anomalies, and the Case of B1555+375: Beware the Effect of Edge-on Disks

### 2015-05-03 papers/B1555_EdgeOn/B1555_letter.tex
Update to Leon's comments

A common feature of simulations of structure formation is the presence
of thousands of subhaloes that are associated with larger mass haloes \citep[e.g.][]{Springel08}.
However, observations of the Local Group find many fewer satellite
galaxies than are predicted by the simulations, even when taking into
account completeness corrections arising from the limited sky coverage
imposed by the Galactic Plane and the latest discovery of new dwarf galaxies \citep{DES15,Kop15}.  
This is the famous ``missing satellite'' problem \citep{Klypin1999, Moore1999, S07}. In order to
understand this discrepancy, it is imperative to build up a large
sample of satellites/subhaloes in galaxies outside the Local Group.
 However, this process is challenging due to the faintness of the satellite
galaxies.  Furthermore, lower-mass satellites may not be able to
retain the gas needed for ongoing star formation \citep[e.g.,][]{P11},
rendering them effectively dark.  These factors make gravitational
lensing a powerful and promising tool for the detection of
substructure.

It was first suggested by \citet{Mao1998} that flux-ratio anomalies
observed in radio-loud multiple images of lensed quasars could be
interpreted as a telltale sign of the presence of substructure in lens
galaxies.  Indeed, small perturbations in the gravitational potential
are sufficient to cause flux-ratio anomalies in strong lensed systems
\citep{Bradac02}.  Moreover, unlike in the optical or infrared, in the
radio band, lensed images are neither sensitive to microlensing from
stars, nor to dust extinction. This makes flux-ratio anomalies a promising
tool to detect \textbf{and measure} dark matter substructures \citep{Dalal2002}.  At
present, the amount of substructure derived from flux-ratio anomalies
is in disagreement with predictions from numerical simulations, with
the former requiring a larger fraction of mass in substructure than
expected \citep{Xu14}. While this could be the result of small sample
statistics, it could also be an indication that flux-ratio anomalies
have a different origin than substructure.
%
An alternative method for
the detection of substructure in gravitational lens galaxies, utilizing
the surface brightness distribution of extended arcs and Einstein rings
has been introduced by \citet{K05} and \citet{V09}. Interestingly, the fraction of
substructure measured with this \emph{gravitational imaging technique}
is currently in agreement with theoretical expectations \citep{V14a,V12}.

%While future large-scale surveys, such as the LSST, are expected to
%deliver new large samples of gravitationally lensed quasars, in the
%near future the discovery of new systems using narrow-line quasar
%emission \citep{N14} will provide new invaluable statistical
%perspective on current constraints based on flux-ratio anomalies.  
In this letter, we explore the idea that not all flux-ratio anomalies are
related to the presence of substructure, but that some originate in a
more complex lens galaxy mass distribution than initially
considered. To this end we use new infra-red data from the Strong
lensing at High Angular Resolution Program 
\citep[SHARP;][Fassnacht et al., in prep]{SHARP12, V12}.
The main goal of the SHARP project is to re-observe known
quadruple and Einstein ring lens systems at high angular
resolution by making use of observations with (1) adaptive optics (AO)
on the Keck telescopes, (2) the Hubble Space Telescope, and (3) large
interferometric arrays \citep{SHARP12,V12}. In particular, 
we focus on the B1555+375 lens system\citep{Marlow99} to better
understand the causes of flux-ratio anomalies in gravitational lenses.  This system was
discovered in the Cosmic Lens All-Sky Survey \citep{CLASS1,CLASS2} and
has a pair of merging images that show a strong flux-ratio anomaly at
radio wavelengths.  Follow-up observations ruled out radio propagation
effects as the cause of the anomaly \citep{K03,KD04}.  The B1555+375
system was included in both the observational and theoretical analysis
of \citet{Dalal2002} and \citet{Xu14} and, thus, understanding its
mass distribution is of key importance.  
%Our analysis makes use of
%the new information obtained from the latest SHARP infrared imaging.
The paper is organized as follows: the  SHARP data are
introduced in section 2; in section 3 we present the lens model while
the results of this analysis and their implications are discussed in
section 4.

@@

\begin{table}
 \centering
 %\begin{minipage}{140mm}
  \caption{Multiband observation details of the B1555+375 lens system.}
  \begin{tabular}{@{}llccc}

@@

To model B1555+375, we use the lens modelling code {\tt glafic}
\citep{Oguri}.  The inputs to the model are the observed image
positions and flux densities measured by the radio observations.
\textbf{We use the image positions from \citet{Marlow99}, who also presented a singular isothermal ellipsoid
(SIE) model that provided a good fit to the image positions but
does not reproduce the strong flux-ratio anomaly between components A
and B.}  %\citep[see Fig.~6 and Tables 2 \& 3 in][]{Marlow99}.  
\textbf{The flux measurements are performed by \citet{K03} with a half year monitoring to obtain flux-ratio curves, providing throughout discussions in erros on flux densities than in \citet{Marlow}.}

Although the flux-ratio anomaly that is observed in the radio data \textbf{implies there must be some perturbation to the mass model, which} may be
caused by substructure in the lensing galaxy, our high-resolution near
infrared imaging suggests another explanation.  In both the $K^\prime$
band adaptive optics data and the \textit{HST} infrared imaging (\textbf{see} in
Fig. 1), the lensing galaxy has a clear edge-on disk component with a
position angle of $\approx 10$ degree.  We therefore model the system
with a disk component in order to see if such a model can explain the
flux-ratio anomaly without the need for substructure.  The edge-on
disk may also explain the system morphology that is observed in the
near infrared imaging.  In particular, neither image B nor image D is
detected in the Keck AO data (Fig.~\ref{fig:merlin}), and both of
these lensed images lie on or close to the edge-on disk.  Thus, strong
extinction in the disk could cause the \textbf{non-}detection of these
images.

As a first trial of modelling, we create a single SIE model to check
the performance of a simple lensing potential. Consistent with
previous studies \citep{Marlow99, Xu14}, our single elliptical
potential model cannot \textbf{reproduce} the flux anomaly in B1555+375.  We then
create two separate models of the lensing potential that both
incorporate the disk.  In the first we treat the lens as a SIE bulge\textbf{/halo}
and an exponential edge-on disk (SIE+expdisk).  In the second model
the disk is represented as a very elongated SIE component (2-SIE). [WHY MODEL BULGE W/ SIE? COMMENT ON THE REASON FOR PICKING MODELS--LEON]
The lens system is red, with a strong jump in flux between the F814W
(roughly I) and F160W (roughly H) bands (Fig.~\ref{fig:multiband}), so
we assign a lens redshift of $z_\ell = 1.0$ and a source redshift of
$z_s =1.5$. \textbf{This assumption on redshifts is needed because some lens model parameters contain distance information. However, our choice of redshifts does not affect the predicted magnifications nor flux-ratios.}

@@

From a theoretical perspective, \citet{Xu14} also find that substructures
cannot explain the observed radio flux anomalies in all the lenses
that were considered. After adding subhaloes from numerical simulations
to these systems, only one out of the eight radio-loud systems
achieves the level of the flux anomalies seen in the observations.  The rest of
them are unlikely (only $1 - 4$ per cent probability) to have \textbf{observed distribution of flux
anomalies} produced only by substructure, including our target B1555+375.
In fact, among those eight radio lenses, B1555+375 is one of the most
anomalous systems, but it is also one of the most challenging systems
to explain by substructures. \citet{Xu14} have concluded that the failure in
predicting flux-ratios is most likely due to oversimplified/improper lens
modelling. They also point out that the SIE component and the external
shear in their model are orthogonal, indicating a possible missing
ingredient in the lens model. Their results are consistent with the SHARP
$K^{\prime}$ band observations, in which an edge-on disk crosses the
radio merging double and which fully accounts for the observed flux-ratio anomaly.  In general, it is well known 
\citep[e.g,][]{Ka91} that lens models based only on the locations of lensed
quasar/AGN images have large degeneracies. Therefore, before
concluding that the flux-ratios in any given lens system are a result
of substructure, it is critical to explore a range of smooth models
that are informed by additional observations or to do a comprehensive
analysis such as that conducted by \citet{Xu14}. Otherwise, the derived
substructure fractions should be considered as upper limits.

>>>>

A common feature of simulations of structure formation is the presence
of thousands of subhaloes that are associated with larger mass haloes \citep[e.g.][]{Springel08}.
However, observations of the Local Group show many fewer satellite
galaxies than are predicted by the simulations, even when taking into
account completeness corrections arising from the limited sky coverage
imposed by the Galactic Plane and the latest discovery of new dwarf galaxies \citep{DES15,Kop15}.  
This is the famous ``missing satellite'' problem \citep{Klypin1999, Moore1999, S07}. In order to
understand this discrepancy, it is imperative to build up a large
sample of satellites/subhaloes in galaxies outside the Local Group.
 However, this process is challenging due to the faintness of the satellite
galaxies.  Furthermore, lower-mass satellites may not be able to
retain the gas needed for ongoing star formation \citep[e.g.,][]{P11},
rendering them effectively dark.  These factors make gravitational
lensing a powerful and promising tool for the detection of
substructure.

It was first suggested by \citet{Mao1998} that flux-ratio anomalies
observed in radio-loud multiple images of lensed quasars could be
interpreted as a telltale sign of the presence of substructure in lens
galaxies.  Indeed, small perturbations in the gravitational potential
are sufficient to cause flux-ratio anomalies in strong lensed systems
\citep{Bradac02}.  Moreover, unlike in the optical or infrared, in the
radio band, lensed images are neither sensitive to microlensing from
stars, nor to dust extinction. This makes flux-ratio anomalies a promising
tool to detect \textbf{and measure} dark matter substructures \citep{Dalal2002,N13} [NEED MORE REF].  
At present, the amount of substructure derived from flux-ratio anomalies
seems to be disagreement with predictions from numerical simulations, with
the former requiring a larger fraction of mass in substructure than
expected \citep{Xu14}. While this could be the result of small sample
statistics, it could also be an indication that flux-ratio anomalies
have a different origin than substructure.[COMMENT ON LINE OF SIGHT EFFECTS?--LEON]
%
An alternative method for
the detection of substructure in gravitational lens galaxies, utilizing
the surface brightness distribution of extended arcs and Einstein rings
has been introduced by \citet{K05} and \citet{V09}. Interestingly, the fraction of
substructure measured with this \emph{gravitational imaging technique}
is currently in agreement with theoretical expectations \citep{V14a,V12}, \textbf{although the number of detections remains small}.

%While future large-scale surveys, such as the LSST, are expected to
%deliver new large samples of gravitationally lensed quasars, in the
%near future the discovery of new systems using narrow-line quasar
%emission \citep{N14} will provide new invaluable statistical
%perspective on current constraints based on flux-ratio anomalies.  
In this letter, we explore the idea that not all flux-ratio anomalies are
related to the presence of substructure, but that some originate in a
more complex lens galaxy mass distribution than initially
considered. To this end we use new infra-red data from the Strong
lensing at High Angular Resolution Program 
\citep[SHARP;][Fassnacht et al., in prep]{SHARP12, V12}.
The main goal of the SHARP project is to re-observe known
quadruple and Einstein ring lens systems at high angular
resolution by making use of observations with (1) adaptive optics (AO)
on the Keck telescopes, (2) the Hubble Space Telescope, and (3) large
interferometric arrays \citep{SHARP12,V12}. In particular, 
we focus on the B1555+375 lens system\citep{Marlow99} to better
understand the causes of flux-ratio anomalies in gravitational lenses.  This system was
discovered in the Cosmic Lens All-Sky Survey \citep{CLASS1,CLASS2} and
has a pair of merging images that show a strong \textbf{persistent} flux-ratio anomaly at
radio wavelengths.  Follow-up observations ruled out radio propagation
effects as the cause of the anomaly \citep{K03,KD04}.  The B1555+375
system was included in both the observational and theoretical analysis
of \citet{Dalal2002} and \citet{Xu14} and, thus, understanding its
mass distribution is of key importance \textbf{to assess systematics in these statistics}.  
%Our analysis makes use of
%the new information obtained from the latest SHARP infrared imaging.
The paper is organized as follows: the  SHARP data are
introduced in section 2; in section 3 we present the lens model while
the results of this analysis and their implications are discussed in
section 4.

@@

\begin{table}
 \centering
 %\begin{minipage}{140mm}
  \caption{Observational details of the B1555+375 lens system.}
  \begin{tabular}{@{}llccc}

@@

To model B1555+375, we use the lens modelling code {\tt glafic}
\citep{Oguri}.  The inputs to the model are the observed image
positions and flux densities measured by the radio observations.
\textbf{We use the image positions from \citet{Marlow99}, who also presented a singular isothermal ellipsoid
(SIE) model that provided a good fit to the image positions but
does not reproduce the strong flux-ratio anomaly between components A
and B.}  %\citep[see Fig.~6 and Tables 2 \& 3 in][]{Marlow99}.  
\textbf{The flux measurements are performed by \citet{K03} with a half year monitoring to obtain flux-ratio curves, providing throughout discussions in erros on flux densities.}

Although the flux-ratio anomaly that is observed in the radio data \textbf{implies there must be some perturbation to the mass model, which} may be
caused by substructure in the lensing galaxy, our high-resolution near
infrared imaging suggests another explanation.  In both the $K^\prime$
band adaptive optics data and the \textit{HST} infrared imaging (\textbf{see} in
Fig. 1), the lensing galaxy has a clear edge-on disk component with a
position angle of $\approx 10$ degree.  We therefore model the system
with a disk component in order to see if such a model can explain the
flux-ratio anomaly without the need for substructure.  The edge-on
disk may also explain the system morphology that is observed in the
near infrared imaging.  In particular, neither image B nor image D is
detected in the Keck AO data (Fig.~\ref{fig:merlin}), and both of
these lensed images lie on or close to the edge-on disk.  Thus, strong
extinction in the disk could cause the \textbf{non-}detection of these
images.[REF B0128 PAPER?--LEON]

As a first trial of modelling, we create a single SIE model to check
the performance of a simple lensing potential. Consistent with
previous studies \citep{Marlow99, Xu14}, a single elliptical
potential model cannot \textbf{reproduce} the flux anomaly in B1555+375.
[WHY MODEL BULGE W/ SIE? COMMENT ON THE REASON FOR PICKING MODELS--LEON]
[I TRY TO EXPLAIN HERE AND MAY NEED TO BE MORE FOCUS--JW]
\textbf{We notice that the lensed images are close to the center of B1555+375 since its edge-on disk spreads further out of the Einstein radius in the \textit{HST} imaging. Instead of constructing the lens galaxy with a primary NFW halo, a secondary bulge, plus an elongated disk, we consider only the bulge plus an disk component, reducing the number of free parameters in the model. Also with only four lensed images, mass profiles that are too complicated will not be informative in the lens modelling.}
We then create two separate models of the lensing potential that both
incorporate the disk.  In the first we treat the lens as a SIE bulge\textbf{/halo}
and an exponential edge-on disk (SIE+expdisk). In the second model
the disk is represented as a very elongated SIE component (2-SIE). 
The lens system is red, with a strong jump in flux between the F814W
(roughly I) and F160W (roughly H) bands (Fig.~\ref{fig:multiband}), so
we assign a lens redshift of $z_\ell = 1.0$ and a source redshift of
$z_s =1.5$. \textbf{This assumption on redshifts is needed because some lens model parameters contain distance information. However, our choice of redshifts does not affect the predicted magnifications nor flux-ratios.}

@@

From a theoretical perspective, \citet{Xu14} also find that substructures
cannot explain the observed radio flux anomalies in all the lenses
that were considered. After adding subhaloes from numerical simulations
to these systems, only one out of the eight radio-loud systems
achieves the level of the flux anomalies seen in the observations.  The rest of
them are unlikely (only $1 - 4$ per cent probability) to have \textbf{observed distribution of flux
anomalies} produced only by substructure, including our target B1555+375.
In fact, among those eight radio lenses, B1555+375 is one of the most
anomalous systems, but it is also one of the most challenging systems
to explain by substructures. \citet{Xu14} have concluded that the failure in
predicting flux-ratios is most likely due to oversimplified/improper lens
modelling. They also point out that the SIE component and the external
shear in their model are orthogonal, indicating a possible missing
ingredient in the lens model.
\textbf{Indeed, it's hard to find physical interpretation for an large [IS $\gamma = 0.14$ A LARGE SHEAR?] and orthogonal external shear.
The SHARP $K^{\prime}$ band observations provide another explaination to their result. The edge-on disk crosses the
radio merging double is consistent of this missing component and fully accounts for the observed flux-ratio anomaly.} 
In general, it is well known  \citep[e.g,][]{Ka91} that lens models based only on the locations of lensed
quasar/AGN images have large degeneracies. Therefore, before
concluding that the flux-ratios in any given lens system are a result
of substructure, it is critical to explore a range of smooth models
that are informed by additional observations or to do a comprehensive
analysis such as that conducted by \citet{Xu14}. Otherwise, the derived
substructure fractions should be considered as upper limits.
